---
title: "Process Census Metadata"
author: "Earl F Glynn<br><small>watchdoglab.substack.com/</small>"
date: "<small>`r Sys.Date()`</small>"
output:
  html_document:
    code_download: true
    theme: cerulean
    toc: yes
    toc_depth:  3
    toc_float:
      collapsed:  yes
      smooth_scroll: yes
    number_sections: yes
    code_folding:  show
---

```{r setup, echo = FALSE}
# http://biostat.mc.vanderbilt.edu/wiki/Main/KnitrHtmlTemplate
require(Hmisc)    # provides knitrSet and other functions
knitrSet(lang = 'markdown',   # If using blogdown: knitrSet(lang='blogdown')
         fig.align = 'left',
         w = 6.5,
         h = 4.5,
         cache = FALSE)
```

`r hidingTOC(buttonLabel = "Outline")`

```{r startYourEngines, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA)

time.1 <- Sys.time()
```

# Setup {.tabset .tabset-fade .tabset-pills}

## {.active}

## Packages

```{r Packages}
library(tidyverse)
library(lubridate)
```

Display formatting

```{r Display}
library(kableExtra)  # kable_styling
#library(IRdisplay)   # display_html (needed in Jupyter notebooks)
```

I/O

```{r IO}
library(readxl)      # read_xlsx
library(readr)       # write_csv
library(writexl)     # write_xlsx
```

 Graphics

```{r Graphics}
library(scales)      # comma, comma_format
```

## Helper functions

```{r Helpers}
Show <- function(data, caption="", bigMark="",
                 height = NULL, width = NULL, ...)
{
  data                                       %>%
  kable("html", caption=caption,
        format.args=list(big.mark=bigMark))  %>%
  kable_styling(bootstrap_options=c("striped", "bordered", "condensed"),
                position="left",
                full_width=FALSE, ...)       %>%
  scroll_box(height = height, width = width) 
}
```

# Census Data

## Raw Data

```{r RawData}
raw <- 
  read_csv("ACSST5Y2022.S0101-Data.csv",
           show_col_types = FALSE)         |>
  slice(-1)                                |>   # remove verbose metadata in row 1
  mutate(NAME = str_replace(NAME, " County, Kansas", ""))

dim(raw)
```

## Metadata

Census Mdatadata

create subfields to understand patterns in metadata

```{r Metadata}
meta <- 
  read_csv("ACSST5Y2022.S0101-Column-Metadata.csv",
           show_col_types = FALSE)                        |>
  separate(`Column Name`, c("Table", "Set", "Variable"),
           sep = "_", remove = FALSE)                     |>
  separate(Variable, c("Number", "Suffix"), sep = 3)      |>
  separate(Label, c("Type", "Grouping", "Product", "Age", "Interval"),
           sep = "!!", remove = FALSE)                    |>
  relocate(Table, Set, Number, Suffix, 
           Type, Grouping, Product, Age, Interval)        |>
  mutate(index = row_number())                            |>
  relocate(index)
  
dim(meta)
```
```{r}
head(meta, 10) |> Show()
```


```{r}
write_xlsx(meta, "Census-Meta-Data-All.xlsx")
```

# Metadata by Sex and Age

```{r}
desired <- 
  meta  |>
  filter(
          is.na(Set)    | Set %in% c("C01", "C03", "C05"),  # raw data, not percentages
          is.na(Suffix) | Suffix == "E",     # estimated 
          is.na(Interval) | Interval %in% 
                          c("Under 5 years",   "5 to 9 years",  "10 to 14 years",
                              "15 to 19 years", "20 to 24 years", "25 to 29 years",
                              "30 to 34 years", "35 to 39 years", "40 to 44 years",
                              "45 to 49 years", "50 to 54 years", "55 to 59 years",
                              "60 to 64 years", "65 to 69 years", "70 to 74 years",
                              "75 to 79 years", "80 to 84 years", "85 years and over",
                              "Under 18 years", "18 years and over")
        ) |>
  mutate(
          varName = case_when(
                              Table == "NAME"                   ~ "GEONAME",
                              Age == "SELECTED AGE CATEGORIES"  ~ paste0("Age_",Interval),
                              Age == "AGE"                      ~ paste0("Age_",Interval),
                              is.na(Interval)                   ~ "Total",         
                              .default =                        "not assigned"
                             ),
         varName = str_replace(varName, " and over", "_and_over") |>
                   str_replace("Under ",    "Under_")    |>
                   str_replace(" to ",      "_")         |>
                   str_replace(" years",    "")          |>
                   str_replace("Under ",    "Under_")    |>
                   str_replace(" and over", "_and_over")
        )

dim(desired)
```

```{r}
write_xlsx(desired, "Census-Meta-Data-Desired.xlsx")
```

# Split into Subsets

# Total 

Select Columns

```{r}
subsetTotalColumns <-
  desired  |>
  filter(is.na(Number) | Set == "C01")   #####

dim(subsetTotalColumns)
```

Read Columns

```{r}
ageWideTotal <-
  raw |>
  select(all_of(subsetTotalColumns$`Column Name`)) |>
  rename_with(~ subsetTotalColumns$varName)        |> # here "~" means "function"
  mutate(across(-GEONAME, as.numeric),
         Set = "Total")                            |>
  relocate(Set)                                    |>
  mutate(Age_18_19 = Age_18_and_over -
                     Age_20_24 - Age_25_29 - Age_30_34 - Age_35_39 - Age_40_44 -
                     Age_45_49 - Age_50_54 - Age_55_59 - Age_60_64 - Age_65_69 -
                     Age_70_74 - Age_75_79 - Age_80_84 - Age_85_and_over)         |>
  relocate(Age_18_19, .after = Age_15_19)

dim(ageWideTotal)
```

```{r}
write_xlsx(ageWideTotal, "Census-Kansas-Total-Wide-S0101-ACS-5yr-2022.xlsx")
```


# Male

Select Columns

```{r}
subsetMaleColumns <-
  desired  |>
  filter(is.na(Number) | Set == "C03")   #####

dim(subsetMaleColumns)
```

Read Columns

```{r}
ageWideMale <-
  raw |>
  select(all_of(subsetMaleColumns$`Column Name`)) |>
  rename_with(~ subsetMaleColumns$varName)        |> # here "~" means "function"
  mutate(across(-GEONAME, as.numeric),
         Set = "Male")                            |>
  relocate(Set)                                   |>
  mutate(Age_18_19 = Age_18_and_over -
                     Age_20_24 - Age_25_29 - Age_30_34 - Age_35_39 - Age_40_44 -
                     Age_45_49 - Age_50_54 - Age_55_59 - Age_60_64 - Age_65_69 -
                     Age_70_74 - Age_75_79 - Age_80_84 - Age_85_and_over)         |>
  relocate(Age_18_19, .after = Age_15_19)

dim(ageWideMale)
```

```{r}
write_xlsx(ageWideMale, "Census-Kansas-Male-Wide-S0101-ACS-5yr-2022.xlsx")
```
                                 
# Female

Select Columns

```{r}
subsetFemaleColumns <-
  desired  |>
  filter(is.na(Number) | Set == "C05")   #####

dim(subsetFemaleColumns)
```

Read Columns

```{r}
ageWideFemale <-
  raw |>
  select(all_of(subsetFemaleColumns$`Column Name`)) |>
  rename_with(~ subsetFemaleColumns$varName)        |> # here "~" means "function"
  mutate(across(-GEONAME, as.numeric),
         Set = "Female")                            |>
  relocate(Set)                                     |>
  mutate(Age_18_19 = Age_18_and_over -
                     Age_20_24 - Age_25_29 - Age_30_34 - Age_35_39 - Age_40_44 -
                     Age_45_49 - Age_50_54 - Age_55_59 - Age_60_64 - Age_65_69 -
                     Age_70_74 - Age_75_79 - Age_80_84 - Age_85_and_over)         |>
  relocate(Age_18_19, .after = Age_15_19)

dim(ageWideFemale)
```

```{r}
write_xlsx(ageWideFemale, "Census-Kansas-Female-Wide-S0101-ACS-5yr-2022.xlsx")
```


# Epilog {.tabset .tabset-fade .tabset-pills}

## {.active}

## Session Info

```{r devtoolsSessionInfo}
devtools::session_info()
```

</div>

```{r epilogDeltaTime, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", sprintf("%.1f",
                        as.numeric(difftime(time.2,
                                            time.1, units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`

