---
title: "Reformat Census"
author: "Earl F Glynn<br><small>watchdoglab.substack.com/</small>"
date: "<small>`r Sys.Date()`</small>"
output:
  html_document:
    code_download: true
    theme: cerulean
    toc: yes
    toc_depth:  3
    toc_float:
      collapsed:  yes
      smooth_scroll: yes
    number_sections: yes
    code_folding:  show
---

```{r setup, echo = FALSE}
# http://biostat.mc.vanderbilt.edu/wiki/Main/KnitrHtmlTemplate
require(Hmisc)    # provides knitrSet and other functions
knitrSet(lang = 'markdown',   # If using blogdown: knitrSet(lang='blogdown')
         fig.align = 'left',
         w = 6.5,
         h = 4.5,
         cache = FALSE)
```

`r hidingTOC(buttonLabel = "Outline")`

```{r startYourEngines, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA)

time.1 <- Sys.time()
```

# Setup {.tabset .tabset-fade .tabset-pills}

## {.active}

## Constants

```{r Constants}
FILE_PREFIX   <- "x"    # no spaces
```

```{r}
EXCEL_LIMIT <- 2^20
```

```{r}
plotCaptionLeft  <- "IRS 527"
plotCaptionRight <- paste("watchdoglab.substack.com", format(Sys.time(), "%Y-%m-%d"))
```

## Packages

```{r Packages}
library(tidyverse)
library(lubridate)
```

Display formatting

```{r Display}
library(kableExtra)  # kable_styling
#library(IRdisplay)   # display_html (needed in Jupyter notebooks)
```

I/O

```{r IO}
library(readxl)      # read_xlsx
library(readr)       # write_csv
library(writexl)     # write_xlsx
```

 Graphics

```{r Graphics}
library(scales)      # comma, comma_format
```

Utilities

```{r Utilities}
library(downloader)  # download
library(tools)       # md5sum
```

## ggplot theme

[ggplot Themes Gallery](https://www.datanovia.com/en/blog/ggplot-themes-gallery/)

```{r ggplotTheme}
theme_set(theme_minimal() +

          theme(axis.text             = element_text(size = 10),
                axis.title            = element_text(size = 14),

                legend.position       = "bottom",

                plot.caption          = element_text(hjust = c(0.0,1.0),
                                                     size = 10),
                plot.caption.position = "plot",

                plot.title.position   = "plot",

                strip.background      = element_rect(fill = "aliceblue"),
                strip.text            = element_text(size = 14),

                title                 = element_text(size = 14)))

COLOR_BAR     <- "skyblue"
COLOR_OUTLINE <- "grey80"
```

## Helper functions

```{r Helpers}
Show <- function(data, caption="", bigMark="",
                 height = NULL, width = NULL, ...)
{
  data                                       %>%
  kable("html", caption=caption,
        format.args=list(big.mark=bigMark))  %>%
  kable_styling(bootstrap_options=c("striped", "bordered", "condensed"),
                position="left",
                full_width=FALSE, ...)       %>%
  scroll_box(height = height, width = width) 
}
```

# Read Census File Headers

```{r StartHere}
meta <- 
  read_csv("ACSST5Y2022.S0101-Column-Metadata.csv",
           show_col_types = FALSE)                        |>
  separate(`Column Name`, c("Table", "Set", "Variable"),
           sep = "_", remove = FALSE)                     |>
  separate(Variable, c("Number", "Suffix"), sep = 3)      |>
  separate(Label, c("Type", "Grouping", "Product", "Age", "Interval"),
           sep = "!!", remove = FALSE)                    |>
  relocate(Table, Set, Number, Suffix, 
           Type, Grouping, Product, Age, Interval)        |>
  mutate(index = row_number())                            |>
  relocate(index)
  
dim(meta)
```
```{r}
head(meta, 10) |> Show()
```


```{r}
write_xlsx(meta, "Census-Meta-Data.xlsx")
```

Set:  C01, C03, C05, ID, blanks
Suffix:  E, blanks
Interval:  under 5 years, 5 to 9, 10 to 14, 15 to 19, 20 to 24, 25 to 29, 30 to 34, 35 to 39, 40 to 44, 
           45 to 49, 50 to 54, 55 to 59, 60 to 64, 45 to 69, 70 to 74, 75 to 79, 80 to 84, 85 and over,
           under 18, 18 and over,
           blanks




# Extract Estimates

```{r}
estimate <- 
  d |>
  select(NAME = `Geographic Area Name`,
         starts_with("Estimate!!Total"),
         starts_with("Estimate!!Male"),
         starts_with("Estimate!!Female"))

dim(estimate)
```

```{r}
glimpse(estimate)
```


# Epilog {.tabset .tabset-fade .tabset-pills}

## {.active}

## Session Info

```{r devtoolsSessionInfo}
devtools::session_info()
```

</div>

```{r epilogDeltaTime, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:", sprintf("%.1f",
                        as.numeric(difftime(time.2,
                                            time.1, units="secs"))), "secs\n")
```

`r processingTime`
`r format(time.2, "%Y-%m-%d %H%M")`

